Likelihoods for a general class of ARGs under the SMC

### 2025-02-21 paper.tex
More updates on the intro

% TODO Maybe swap the order of these two paragraphs around, so it's 
% more natrual to talk about the advantages of model-driven stuff next.
The first class of inference methods are Bayesian approaches
that sample ARGs under a population genetics model such as the
coalescent with recombination~\citep{hudson_properties_1983},
and its approximatation, the Sequentially Markovian
Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
ARGweaver~\citep{rasmussen_genome-wide_2014} has been the most widely used and studied,
and until recently consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
and has succesfully been applied in several
contexts~\citep[e.g.,][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}. 
The recently-introduced method SINGER takes a similar approach,
and by using an improved Monte Carlo algorithm
promises to be even more accurate and substantially more 
scalable~\citep{deng_robust_2024}. 
The second class of inference methods that has been of recent interest 
are based on heuristics and approximate models.
Relate~\citep{speidel_method_2019},
tsinfer~\citep{kelleher_inferring_2019},
ARG-Needle~\citep{zhang_biobank-scale_2023} 
and Threads~\citep{gunnarsson_scalable_2024}
work on quite different 
principles, but share some common properties. 
Firstly, they are all 
heuristic and approximation driven, favouring computational
scalability over a basis in a statistical model.
Secondly, they all produce a single
ARG as output, inferring a deterministic point estimate.
Thirdly, they regard the precise timing of nodes as a separate
problem, focusing only on the relative ordering of nodes when 
producing the initial ARG. General purpose ARG dating methods are 
appearing~\citep{wohns_unified_2022,deng2025general} and can also improve
dating performance in ARGs sampled from the SMC~\citep{deng2025general}.
Fourthly (and most importantly for the 
purposes of this paper) the ARGs that hueristic methods produce 
lack some of the topological information present in the output of ARGweaver or SINGER. 
This occurs, roughly speaking, because of a basic difference in their
approach to uncertainty.

There is considerable uncertainty in ARG inference, and often a fundamental
lack of information in the sequence data to distinguish different
possibilities. Consider, for example, a case in which we have three
ancestral lineages with identical sequences. We know that there
must have been two coalescence events, but there is no mutational information to 
distinguish their relative ordering. Sampling methods overcome this problem
by randomly choosing an ordering, and uncertainty communicated by the order
changing in the different ARGs sampled from the posterior.
The other approach (used by methods such as tsinfer) is to communicate
the uncertainty structurally by means of a polytomy: we have no
information about the intermediate coalescent event, and so we omit it.
Similarly, there is often a fundamental lack of information about the 
details of recombination events, and heuristic methods (in different ways)
omit these details. In many ways, the significant progress made in
scalability by these heuristic methods is \emph{because} they don't
attempt to precisely reconstruct recombination events.

As there is no mutational
information

(e.g., the timing of recombination events, as discussed below).

and the heuristic methods deal with this uncertainty by omitting these events,
while ARGweaver and SINGER provide a sample from the posterior.
(So, some aspects of the resulting ARGs should be interpreted slightly differently;
see \citet{wong_general_2023} for more discussion.)

There are substantial advantages to the statistically rigorous sampling 
approaches of ARGweaver and SINGER.
Firstly, they tend to be more accurate when compared on the same data set.
Secondly, the direct connection with population genetic models allows for
natural extensions like the joint estimation of ARGs and
demography~\citep{hubisz_mapping_2020}.
Thirdly, sampling ARGs from a
posterior distribution provides a means of quantifying the uncertainty 
around estimates, which can be used in downstream applications.

On the other hand, heuristic methods have the potential
to utilize the greater amount of information in large datasets,
thus possibly producing even more accurate estimates
(although this has not, to our knowledge, been tested).
Although SINGER is a major step forward
in terms of scalability over ARGweaver and other methods, inference is still 
only feasible for hundreds of samples. 
Meanwhile, tsinfer, ARG-Needle and Threads have 
been applied to datasets of hundreds of \emph{thousands} of samples.
With the explosive growth in dataset
sizes in recent years~\citep{caulfield2017national,turnbull2018100,
bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole,
all2024genomic,stark2024call,cook2025our}
there is a pressing need for more statistical
rigor in large scale ARG inference.
% This inability to connect the results of large-scale inference methods
% with a generative model
% is a fundamental limitation, and limits both the utility of 
% current point estimates as well as the possibilities for further
% inference. Specifically, we are unable to quantify uncertainty 
% around heuristic point estimates. We cannot explore ARG space
% around the point estimates using (e.g., MCMC), as we have no way definition
% of improvement. We cannot compare models of demography, or evaluate the 
% goodness of fit for model estimates.
A key problem facing the field is that the ARGs estimated by heuristic large-scale 
methods lack a concrete connection to population genetic theory because
we cannot compute their likelihood under the SMC or similar models. 
% The classical method~\citep{kuhner_maximum_2000} assumes complete knowledge about
% recombination events, and there is no clear way to compute a likelihood for 
% the approximate structures output by heuristic methods.

Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner-Yamato-Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probablility to each recombination and common
ancestor event, and to the inter-event  waiting times. 
The likelihood of an 
ARG is then a product over all these three classes of term.
Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG given a complete description of

, such complete detail is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical ms
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail for likelihood computation.
The msprime
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to msprime~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} similarly do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.

>>>>

% TODO Maybe swap the order of these two paragraphs around, so it's 
% more natrual to talk about the advantages of model-driven stuff next.
The first class of inference methods are Bayesian approaches
that sample ARGs under a population genetics model such as the
coalescent with recombination~\citep{hudson_properties_1983},
and its approximatation, the Sequentially Markovian
Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
ARGweaver~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020} 
has been the most widely used and studied,
and until recently consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
and has succesfully been applied in several
contexts~\citep[e.g.,][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}. 
The recently-introduced method SINGER takes a similar approach,
and by using an improved Monte Carlo algorithm
promises to be even more accurate and substantially more 
scalable~\citep{deng_robust_2024}. 
The second class of inference methods that has been of recent interest 
are based on heuristics and approximate models.
Relate~\citep{speidel_method_2019},
tsinfer~\citep{kelleher_inferring_2019},
ARG-Needle~\citep{zhang_biobank-scale_2023} 
and Threads~\citep{gunnarsson_scalable_2024}
work on quite different 
principles, but share some common properties. 
Firstly, they are all 
heuristic and approximation driven, favouring computational
scalability over a basis in a statistical model.
Secondly, they all produce a single
ARG as output, inferring a deterministic point estimate.
Thirdly, they regard the precise timing of nodes as a separate
problem, focusing only on the relative ordering of nodes when 
producing the initial ARG. General purpose ARG dating methods are 
appearing~\citep{wohns_unified_2022,deng2025general} and can also improve
dating performance in ARGs sampled from the SMC~\citep{deng2025general}.
Fourthly (and most importantly for the 
purposes of this paper) the ARGs that hueristic methods produce 
lack some of the topological information present in the output of ARGweaver or SINGER. 
This occurs, roughly speaking, because of a basic difference in their
approach to uncertainty.

There is considerable uncertainty in ARG inference, and often a fundamental
lack of information in the sequence data to distinguish different
possibilities. Consider, for example, a case in which we have three
ancestral lineages with identical sequences. We know that there
must have been two coalescence events, but there is no mutational information to 
distinguish their relative ordering. Sampling methods overcome this problem
by randomly choosing an ordering, with the uncertainty communicated by the order
changing in the different ARGs sampled from the posterior.
The other approach (used by methods such as tsinfer) is to communicate
the uncertainty structurally by means of a polytomy: we have no
information about the intermediate coalescent event, and so we omit it.
Similarly, there is often a fundamental lack of information about the 
details of recombination events, and heuristic methods (in different ways)
omit these details. In many ways, the significant progress made in
scalability by these heuristic methods is \emph{because} they don't
attempt to precisely reconstruct recombination events
(this should be taken into account when interpreting the resulting ARGs;
see \citet{wong_general_2023} for more discussion).

There are substantial advantages to the statistically rigorous sampling 
approaches of ARGweaver and SINGER.
In particular, they tend to be more accurate (at the 
scale at which they can be compared),
and sampling ARGs from a
posterior distribution provides a means of quantifying the uncertainty 
around estimates.
On the other hand, heuristic methods have the potential
to use the greater amount of information in large datasets,
capturing fine-scale details about the recent past.
% thus possibly producing even more accurate estimates
% (although this has not, to our knowledge, been tested).
Although SINGER is a major step forward
in terms of scalability over ARGweaver and other methods, inference is still 
only feasible for hundreds of samples. 
Meanwhile, tsinfer, ARG-Needle and Threads have 
been applied to datasets of hundreds of \emph{thousands} of samples.
With the explosive growth in dataset
sizes in recent years~\citep{caulfield2017national,turnbull2018100,
bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole,
all2024genomic,stark2024call,cook2025our}
there is a pressing need for more statistical
rigor in large scale ARG inference.
% This inability to connect the results of large-scale inference methods
% with a generative model
% is a fundamental limitation, and limits both the utility of 
% current point estimates as well as the possibilities for further
% inference. Specifically, we are unable to quantify uncertainty 
% around heuristic point estimates. We cannot explore ARG space
% around the point estimates using (e.g., MCMC), as we have no way definition
% of improvement. We cannot compare models of demography, or evaluate the 
% goodness of fit for model estimates.

A key problem facing the field is that the ARGs estimated by heuristic large-scale 
methods lack a concrete connection to population genetic theory because
we cannot compute their likelihood under the SMC or similar models. 
Computing the likelihood of an ARG under a model is a fundamental element of
probabilistic inference. The classical Kuhner-Yamato-Felsenstein
\citeyearpar{kuhner_maximum_2000} formulation (henceforth: KYF)
works by assigning a probablility to each recombination and common
ancestor event, and to the inter-event  waiting times. 
Although the KYF formulation is a natural and elegant way to describe 
the likelihood of an ARG, the high-level of detail about the underlying
events that is required is not present 
in many estimated ARGs~\citep{wong_general_2023} or even in most simulations.
Simulators such as the classical ms
program~\citep{hudson_generating_2002} output ARGs in a tree-by-tree format,
which does not contain sufficient detail for likelihood computation.
The msprime
simulator~\citep{kelleher_efficient_2016,baumdicker_efficient_2021} 
by default outputs a more complete and efficient representation of 
the simulated ancestry, but still does not provide the exhaustive
detail required.
Indeed, providing the information required to 
support likelihood calculations originally motivated the addition of
the \texttt{record\_full\_arg}
option to msprime~\citep{baumdicker_efficient_2021}.
Forwards-time simulations that output 
ARGs~\citep{kelleher_efficient_2018,haller_tree_2018} also do not
typically contain the level of detail required to compute a likelihood under
the KYF approach.
